\documentstyle[portrait,%
              fancybox,%
              notes,%
              epsfig,%
              alltt,%
              semcolor,
              alltt]{seminar}

\input amssym.def
\input amssym

\newcommand{\nat}{\mbox{$\protect\Bbb N$}}
\newcommand{\num}{\mbox{$\protect\Bbb Z$}}
\newcommand{\rat}{\mbox{$\protect\Bbb Q$}}
\newcommand{\real}{\mbox{$\protect\Bbb R$}}
\newcommand{\complex}{\mbox{$\protect\Bbb C$}}
\newcommand{\xxx}{\mbox{$\protect\Bbb X$}}


\newcommand{\True}{\top}
\newcommand{\False}{\bot}
\newcommand{\Not}{\neg}
\newcommand{\And}{\wedge}
\newcommand{\Or}{\vee}
\newcommand{\Imp}{\Rightarrow}
\newcommand{\Iff}{\Leftrightarrow}

\newcommand{\lamb}[1]{\lambda #1.\:}
\newcommand{\eps}[1]{\varepsilon #1.\:}
\newcommand{\all}[1]{\forall #1.\:}
\newcommand{\ex}[1]{\exists #1.\:}
\newcommand{\exu}[1]{\exists! #1.\:}

\newcommand{\BA}{\begin{array}[t]{l}}
\newcommand{\EA}{\end{array}}

\newcommand\leb{\lbrack\!\lbrack}
\newcommand\reb{\rbrack\!\rbrack}
\newcommand{\sem}[1]{\leb #1 \reb}

% Sizes

\renewcommand{\slidetopmargin}{0.8in}
\renewcommand{\slidebottommargin}{0.8in}

% Various combinations of one colour on another

\newcommand{\greenonred}[1]%
{\psset{fillcolor=red}\psframebox*[framearc=.3]{\green #1}}

\newcommand{\whiteonred}[1]%
{\psset{fillcolor=red}\psframebox*[framearc=.3]{\white #1}}

\newcommand{\yellowonmagenta}[1]%
{\psset{fillcolor=magenta}\psframebox*[framearc=.3]{\yellow #1}}

\newcommand{\whiteonblack}[1]%
{\psset{fillcolor=black}\psframebox*[framearc=.3]{\white #1}}

\newcommand{\blackonlightgray}[1]%
{\psset{fillcolor=lightgray}\psframebox*[framearc=.3]{\black #1}}

\newcommand{\blueonlightgray}[1]%
{\psset{fillcolor=lightgray}\psframebox*[framearc=.3]{\blue #1}}

\newcommand{\cyanonblack}[1]%
{\psset{fillcolor=black}\psframebox*[framearc=.3]{\cyan #1}}

\newcommand{\blueonyellow}[1]%
{\psset{fillcolor=yellow}\psframebox*[framearc=.3]{\blue #1}}

\newcommand{\redonyellow}[1]%
{\psset{fillcolor=yellow}\psframebox*[framearc=.3]{\red #1}}

\newcommand{\heading}[1]%
{\begin{center}\whiteonblack{\large\bf\blueonlightgray{#1}}\end{center}}

\newcommand{\emphatic}[1]{\blueonyellow{#1}}

\newcommand{\veryemphatic}[1]%
{\begin{center}{\emphatic{#1}}\end{center}}

% Head and foot of slides

\newpagestyle{ColourDemo}%
  {\cyanonblack{Introduction to Functional Programming:
                Lecture 1}\hfil\cyanonblack{\thepage}}
  {\cyanonblack{John Harrison}\hfil
   \cyanonblack{University of Cambridge, 16 January 1997}}

\pagestyle{ColourDemo}

\centerslidesfalse

% Colour bullets

\def\labelitemi{{\black$\bullet$}}
\def\labelitemii{{\black--}}
\def\labelitemiii{{\black$\ast$}}
\def\labelitemiv{{\black$\cdot$}}

% Start of document (default text colour is blue)

\begin{document}\blue


\begin{slide*}

\heading{%
\begin{tabular}{c}
{\LARGE\red Introduction to}\\
{\LARGE\red Functional Programming}\\
{\cyan John Harrison}\\
{\cyan University of Cambridge}\\
{\green Lecture 1}\\
{\green Introduction and Overview}
\end{tabular}}

\vspace*{0.5cm}

Topics covered:

\begin{itemize}

\item Imperative programming

\item Functional programming

\item The merits of functional programming

\item Overview of the course

\item Lambda notation and its benefits

\end{itemize}

\end{slide*}




\begin{slide*}

\heading{Imperative programming}

\vspace*{0.5cm}

Imperative (or procedural) programs rely on modifying a {\em state} by using a
sequence of {\em commands}.

The state is mainly modified by the {\em assignment} command, written {\red \tt
v = E} or {\red \tt v := E}.

We can execute one command before another by writing them in sequence, perhaps
separated by a semicolon: {\red $C_1 \mbox{ ; } C_2$}.

Commands can be executed conditionally using {\red \tt if}, and repeatedly
using {\red \tt while}.

Programs are a series of instructions on how to modify the state.

Imperative languages, e.g. {\black FORTRAN}, {\black Algol}, {\black C},
{\black Modula-3} support this style of programming.

\end{slide*}


\begin{slide*}

\heading{An abstract view}

\vspace*{0.5cm}

We ignore input-output operations, and assume that a program runs for a limited
time, producing a result.

We can consider the execution in an abstract way as:

{\red $$ \sigma_0 \to \sigma_1 \to \sigma_2 \to \cdots \to \sigma_n $$}

The program is started with the computer in an initial state {\red $\sigma_0$},
including the inputs to the program.

The program finishes with the computer in a final state {\red $\sigma_n$},
containing the output(s) of the program.

The state passes through a finite sequence of changes to get from {\red
$\sigma_0$} to {\red $\sigma_n$}; in general, each command may modify the
state.

\end{slide*}



\begin{slide*}

\heading{Functional programming}

\vspace*{0.2cm}

A functional program is simply an {\em expression}, and executing the program
means {\em evaluating} the expression.

\begin{itemize}

\item There is no state, i.e. there are {\red no variables}.

\item Therefore there is {\red no assignment}, since there's nothing to assign
to.

\item And there is {\red no sequencing} and {\red no repetition}, since one
expression does not affect another.

\end{itemize}

But on the positive side:

\begin{itemize}

\item We can have {\red recursive functions}, giving something comparable to
repetition.

\item Functions can be used much more flexibly, e.g. we can have {\red higher
order functions}.

\end{itemize}

Functional languages support this style of programming.


\end{slide*}


\begin{slide*}

\heading{Example: the factorial}

\vspace*{0.5cm}

The factorial function can be written imperatively in C as follows:

\begin{red}\begin{verbatim}
  int fact(int n)
  { int x = 1;
    while (n > 0)
     { x = x * n;
       n = n - 1;
     }
    return x;
  }
\end{verbatim}\end{red}

\noindent whereas it would be expressed in ML as a recursive function:

\begin{red}\begin{verbatim}
  let rec fact n =
    if n = 0 then 1
    else n * fact(n - 1);;
\end{verbatim}\end{red}

\end{slide*}



\begin{slide*}

\heading{Why?}

\vspace*{0.5cm}

At first sight a language without variables, assignment and sequencing looks
very impractical.

We will show in this course how a lot of interesting programming can be done in
the functional style.

Imperative programming languages have arisen as an abstraction of the hardware,
from machine code, through assemblers and macro assemblers, to FORTRAN and
beyond.

Perhaps this is the wrong approach and we should approach the task from the
human side. Maybe functional languages are better suited to people.

But what concrete reasons are there for preferring functional languages?

\end{slide*}



\begin{slide*}

\heading{Merits of functional programming}

\vspace*{1.0cm}

By avoiding variables and assignments, we gain the following advantages:

\begin{itemize}

\item Clearer semantics. Programs correspond more directly to abstract
mathematical objects.

\item More freedom in implementation, e.g. parallelizability.

\end{itemize}

By the more flexible use of functions, we gain:

\begin{itemize}

\item Conciseness and elegance.

\item Better parametrization and modularity of programs.

\item Convenient ways of representing infinite data.

\end{itemize}

\end{slide*}


\begin{slide*}

\heading{Denotational semantics}

\vspace*{0.5cm}

We can identify our ML factorial function with an abstract mathematical
(partial) function {\red $\num \to \num$}:

\begin{red}
$$ \sem{\mbox{fact}}(n) = \left\{ \begin{array}{ll}
                                 n! & \mbox{if $n \geq 0$} \\
                                 \bot & \mbox{otherwise}
                          \end{array} \right. $$
\end{red}

\noindent where {\red $\bot$} denotes undefinedness, since for negative
arguments, the program fails to terminate.

Once we have a state, this simple interpretation no longer works. Here is a C
`function' that doesn't correspond to any mathematical function:

\begin{red}\begin{verbatim}
  int rand(void)
  { static int n = 0;
    return n = 2147001325 * n + 715136305;
  }
\end{verbatim}\end{red}

This gives different results on successive calls!

\end{slide*}


\begin{slide*}

\heading{Semantics of imperative programs}

\vspace*{0.5cm}

In order to give a corresponding semantics to imperative programs, we need to
make the state explicit. For example we can model commands as:

\begin{itemize}

\item Partial functions {\red $\Sigma \to \Sigma$} (Strachey)

\item Relations on {\red $\Sigma \times \Sigma$} (Hoare)

\item Predicate transformers, i.e. total functions {\red $(\Sigma \to bool) \to
(\Sigma \to bool)$} (Dijkstra)

\end{itemize}

If we allow the {\red \tt goto} statement, even these are not enough, and we
need a semantics based on {\em continuations} (Wadsworth, Morris).

All these methods are quite complicated.

With functional programs, we have a real chance of proving their correctness,
or the correctness of certain transformations or optimizations.

\end{slide*}


\begin{slide*}

\heading{Problems with functional programs}

\vspace*{0.5cm}

Functional programming is not without its deficiencies. Some things are harder
to fit into a purely functional model, e.g.

\begin{itemize}

\item Input-output

\item Interactive or continuously running programs (e.g. editors, process
controllers).

\end{itemize}

However, in many ways, infinite data structures can be used to accommodate
these things.

Functional languages also correspond less closely to current hardware, so they
can be less efficient, and it can be hard to reason about their time and space
usage.

ML is not a pure functional language, so you can use variables and assignments
if required. However most of our work is in the pure functional subset.

\end{slide*}



\begin{slide*}

\heading{Overview of the course (1)}

\vspace*{0.5cm}

We start with the theoretical underpinning, $\lambda$-calculus, and move up to
the practical business of programming in ML.

Roughly $1 \over 3$ of the course is devoted to the theoretical parts, and $2
\over 3$ to the practical parts.

The theoretical parts are:

\begin{enumerate}

\item Introduction and overview (this lecture)

\item $\lambda$-calculus as a formal system

\item $\lambda$-calculus as a programming language

\item Types

\end{enumerate}

\end{slide*}


\begin{slide*}

\heading{Overview of the course (2)}

\vspace*{0.5cm}

The practical or `applied' parts of the course are:

\begin{enumerate}

\item Introduction to ML

\item More about polymorphism

\item Recursive types

\item Proofs about programs

\item Further ML

\item Other styles of functional programming

\item ML examples (1)

\item ML examples (2)

\end{enumerate}

\end{slide*}


\begin{slide*}

\vspace*{0.5cm}

\heading{Lambda notation}

Lambda notation is a way of denoting functions, invented by Alonzo Church in
the 1930s. We write:

{\red $$ \lamb{x} E[x] $$ }

to denote `the function of {\red $x$} that yields {\red $E[x]$}'. Here {\red
$E[x]$} is any expression which may or may not contain {\red $x$}.

For example {\red $\lamb{x} x$} is the identity function, which simply returns
its argument, while {\red $\lamb{x} x^2$} is the squaring function.

The letter {\red $\lambda$} is arbitrary, a historical accident. Originally
Church used {\red $\hat{x}.\; E[x]$}, and a series of typesetting errors
transformed it into the present form. One also sees {\red $x \mapsto E[x]$}
and {\red $[x]\; E[x]$}.

\end{slide*}


\begin{slide*}

\vspace*{1.0cm}

\heading{Why?}

In informal mathematics, when talking about a function, one normally gives it a
name: `define {\red $f(x) = E[x]$} \ldots then {\red $\cdots f \cdots$}'.

Similarly, most programming languages will only let you define functions if you
are prepared to give them a name.

This approach is rather inconsistent. It means we are not treating functions on
a par with other mathematical objects. We don't say:

\begin{quote}
Define {\red $x$} and {\red $y$} by {\red $x = 2$} and {\red $y = 4$}
respectively. Then {\red $x x = y$}
\end{quote}

Lambda notation helps to put functions on an equal footing. This is important
in functional programming.

\end{slide*}


\begin{slide*}

\vspace*{0.5cm}

\heading{Benefits of $\lambda$-notation}

The $\lambda$-notation can be useful too in order to make mathematics clearer.

When we discuss {\red $x + x y$}, we are often vague about whether we mean the
value for a particular {\red $x$}, or a function of {\red $x$} (or {\red $y$}
\ldots). Using $\lambda$-notation this can be made explicit.

In fact, using just:

\begin{itemize}

\item Variables, e.g. {\red $x$}, {\red $y$}.

\item Constants, e.g. {\red $3$}, {\red $true$}, {\red $+$}.

\item Applications of functions to arguments, e.g. {\red $f\; x$}.

\item Lambda abstractions of expressions over variables, e.g. {\red $\lamb{x} x
+ y$}

\end{itemize}

we reach a general `abstract syntax' for mathematics.

\end{slide*}



\begin{slide*}

\vspace*{0.5cm}

\heading{Currying}

We can have functions of more than one argument by using, for example
{\red $\lamb{x} \lamb{y} x + y$}.

We can think of this as a function {\red $\num \to (\num \to \num)$}. When
given one argument, say {\red $2$}, it returns a function {\red $\num \to
\num$} that adds {\red $2$} to its argument, and this takes the second
argument. This device is known as {\em currying}, after Haskell Curry.

We adopt the conventions that {\red $\lamb{x\; y} E[x,y]$} means {\red
$\lamb{x} \lamb{y} E[x,y]$}, and that function application associates to the
left, e.g. that {\red $f\; x\; y$} means {\red $(f(x))(y)$}. This supports the
use of currying, e.g.

{\red $$ (\lamb{x\; y} x + y)\;1\;2 = (\lamb{y} 1 + y)\;2 = 1 + 2 $$}

Note that the brackets round function arguments are optional.

\end{slide*}


\begin{slide*}

\vspace*{0.5cm}

\heading{Variable binding}

Many constructs in mathematics {\em bind} variables. The variable {\red $x$} in

{\red $$ \int_0^a 3 x^2 \; dx $$}

is bound, whereas {\red $a$} is free; likewise {\red $n$} is bound and {\red
$k$} free in

{\red $$ \Sigma_{n=0}^k n^2 $$}

Free and bound variables can be quite complicated, as we shall see. Using
$\lambda$-notation, all these variable binding constructs can be broken down so
that the only binding operation is $\lambda$-abstraction.

\end{slide*}



\begin{slide*}

\vspace*{0.5cm}

\heading{Example of differentiation}

For example, the everyday construct {\red ${d \over dx} x^2$} can be analyzed
as follows:

{\red $$ D\; (\lamb{x} \mbox{EXP } x\; 2)\; x $$}

where {\red $D$} is the (curried) differentiation operator {\red $(\real \to
\real) \to \real \to \real$} giving the derivative of its first argument at the
point indicated by its second argument, and {\red $\mbox{EXP}$} the
exponentiation function.

So there are really two instances of {\red $x$} here, one bound, one free. The
same occurs in

{\red $$ \int_0^x 2 x \; dx $$}

but here the everyday notation separates the two properly.

\end{slide*}





\begin{slide*}

\vspace*{0.5cm}

\heading{Russell's paradox}

Originally, Church hoped to use $\lambda$-notation to give a foundation for
mathematics, by introducing constants for logical operations, e.g. {\red
$\Not$} to mean `not'. Unfortunately this turned out to be inconsistent,
because defining {\red $R = \lamb{x} \Not\; (x\; x)$} we have:

{\red $$ R\; R = (\lamb{x} \Not\; (x\; x)) \; R = \Not\; (R\; R) $$}

We know there is a correspondence between sets and their characteristic
function, so if we think of {\red $s\;x$} as meaning {\red $x \in s$}, this is
simply the well known Russell paradox of the set of all sets that do not
contain themselves:

{\red $$ R = \{x \mid x \not\in x\} $$}

\end{slide*}



\end{document}

%%% Local Variables:
%%% TeX-master: "all"
%%% End:
