\documentclass[portrait,%
              fancybox,%
              notes,%
              epsfig,%
              alltt,%
              semcolor,
              alltt]{seminar}

\usepackage[koi8-r]{inputenc}
\usepackage[russian]{babel}

\input amssym.def
\input amssym

\newcommand{\nat}{\mbox{$\protect\Bbb N$}}
\newcommand{\num}{\mbox{$\protect\Bbb Z$}}
\newcommand{\rat}{\mbox{$\protect\Bbb Q$}}
\newcommand{\real}{\mbox{$\protect\Bbb R$}}
\newcommand{\complex}{\mbox{$\protect\Bbb C$}}
\newcommand{\xxx}{\mbox{$\protect\Bbb X$}}

\newcommand{\lamb}[1]{\lambda #1.\:}
\newcommand{\eps}[1]{\varepsilon #1.\:}
\newcommand{\all}[1]{\forall #1.\:}
\newcommand{\ex}[1]{\exists #1.\:}
\newcommand{\exu}[1]{\exists! #1.\:}

\newcommand{\True}{\top}
\newcommand{\False}{\bot}
\newcommand{\Not}{\neg}
\newcommand{\And}{\wedge}
\newcommand{\Or}{\vee}
\newcommand{\Imp}{\Rightarrow}
\newcommand{\Iff}{\Leftrightarrow}

\newcommand{\entails}{\vdash}
\newcommand{\proves}{\vdash}

\newcommand{\Ands}{\bigwedge}
\newcommand{\Ors}{\bigvee}

\newcommand{\BQ}{\mbox{$\ulcorner$}}
\newcommand{\BEQ}{\mbox{\raise4pt\hbox{$\ulcorner$}}}
\newcommand{\EQ}{\mbox{$\urcorner$}}
\newcommand{\EEQ}{\mbox{\raise4pt\hbox{$\urcorner$}}}

\newcommand{\QUOTE}[1]{\mbox{$\BQ #1 \EQ$}}
\let\psubset=\subset                    % Pure TeX: thanks to MAJ %
\let\subset=\subseteq

\newcommand{\powerset}{\wp}             % This is pretty dire...  %

\newcommand{\Union}{\cup}
\newcommand{\Inter}{\cap}
\newcommand{\Unions}{\bigcup}
\newcommand{\Inters}{\bigcap}

\newcommand{\proof}{{\bf \noindent Proof:\ }}
\newcommand{\qed}{Q.E.D.}

\newcommand{\Rule}{\infer}

\newcommand{\restrict}{\upharpoonright} % This is lousy and must be fixed! %

\newcommand{\bigsqcap}{\mbox{\Large{$\sqcap$}}}

\newcommand\leb{\lbrack\!\lbrack}
\newcommand\reb{\rbrack\!\rbrack}
\newcommand{\sem}[1]{\leb #1 \reb}

\newcommand{\BA}{\begin{array}[t]{l}}
\newcommand{\EA}{\end{array}}
\newcommand{\sqle}{\sqsubseteq}
\newcommand{\sqlt}{\sqsubset}

\newcommand{\too}{\twoheadrightarrow}

\newcommand{\Los}{{\L}o{\'s}}

% These are from Mike's notes

\def\alphas{\mathrel{\mathop{\longrightarrow}\limits_{\alpha}}}
\def\betas{\mathrel{\mathop{\longrightarrow}\limits_{\beta}}}
\def\etas{\mathrel{\mathop{\longrightarrow}\limits_{\eta}}}

\def\goesto{\longrightarrow}

\newcommand{\defeq}{\stackrel{\bigtriangleup}{=}}

% Sizes

\newenvironment{rslide}
{\begin{slide*}[240mm,165mm] }
{\end{slide*}}

\renewcommand{\slidetopmargin}{0.8in}
\renewcommand{\slidebottommargin}{0.8in}

% Various combinations of one colour on another

\newcommand{\greenonred}[1]%
{\psset{fillcolor=red}\psframebox*[framearc=.3]{\green #1}}

\newcommand{\whiteonred}[1]%
{\psset{fillcolor=red}\psframebox*[framearc=.3]{\white #1}}

\newcommand{\yellowonmagenta}[1]%
{\psset{fillcolor=magenta}\psframebox*[framearc=.3]{\yellow #1}}

\newcommand{\whiteonblack}[1]%
{\psset{fillcolor=black}\psframebox*[framearc=.3]{\white #1}}

\newcommand{\blackonlightgray}[1]%
{\psset{fillcolor=lightgray}\psframebox*[framearc=.3]{\black #1}}

\newcommand{\blueonlightgray}[1]%
{\psset{fillcolor=lightgray}\psframebox*[framearc=.3]{\blue #1}}

\newcommand{\cyanonblack}[1]%
{\psset{fillcolor=black}\psframebox*[framearc=.3]{\cyan #1}}

\newcommand{\blueonyellow}[1]%
{\psset{fillcolor=yellow}\psframebox*[framearc=.3]{\blue #1}}

\newcommand{\redonyellow}[1]%
{\psset{fillcolor=yellow}\psframebox*[framearc=.3]{\red #1}}

\newcommand{\heading}[1]%
{\begin{center}\whiteonblack{\large\bf\blueonlightgray{#1}}\end{center}}

\newcommand{\emphatic}[1]{\blueonyellow{#1}}

\newcommand{\veryemphatic}[1]%
{\begin{center}{\emphatic{#1}}\end{center}}

% Head and foot of slides

\newpagestyle{ColourDemo}%
  {\cyanonblack{Введение в Функциональное Программирование:
                Лекция 11}\hfil\cyanonblack{\thepage}}
  {\cyanonblack{Джон Харрисон}\hfil
   \cyanonblack{Университет Кембриджа, 7 февраля 1997}}
%\newpagestyle{ColourDemo}%
%  {\cyanonblack{Introduction to Functional Programming:
%                Lecture 11}\hfil\cyanonblack{\thepage}}
%  {\cyanonblack{John Harrison}\hfil
%   \cyanonblack{University of Cambridge, 7 February 1997}}

\pagestyle{ColourDemo}

\centerslidesfalse

% Colour bullets

\def\labelitemi{{\black$\bullet$}}
\def\labelitemii{{\black--}}
\def\labelitemiii{{\black$\ast$}}
\def\labelitemiv{{\black$\cdot$}}

% Start of document (default text colour is blue)

\begin{document}\blue


\begin{rslide}

\heading{%
\begin{tabular}{c}
{\Large\red Введение в}\\
{\Large\red Функциональное}\\
{\Large\red Программирование}\\
{\cyan Джон Харрисон}\\
{\cyan Университет Кембриджа}\\
{\green Лекция 11}\\
{\green Примеры на ML III:}\\
{\green Exact Real Arithmetic}
%{\LARGE\red Introduction to}\\
%{\LARGE\red Functional Programming}\\
%{\cyan John Harrison}\\
%{\cyan University of Cambridge}\\
%{\green Lecture 11}\\
%{\green ML examples III:}\\
%{\green Exact Real Arithmetic}
\end{tabular}}

\vspace*{0.5cm}

Topics covered:

\begin{itemize}

\item Finite representations

\item Real numbers as approximating functions

\item Arbitrary precision integers

\item Operations over reals

\item Caching

\end{itemize}

\end{rslide}




\begin{rslide}

\heading{Finite representations}

\vspace*{0.5cm}

Real arithmetic on computers is normally done via floating point
approximations.

In general, we can only manipulate a real number, either ourselves or inside a
computer, via some sort of finite representation.

Some question how numbers can be said to `exist' if they have no finite
representation.

For example, Kronecker accepted integers and rationals because
they can be written down explicitly, and even {\em algebraic}
numbers because they can be represented using the polynomials of which they
are solutions.

However he rejected transcendental numbers because apparently
they could not be represented finitely.

\end{rslide}


\begin{rslide}

\heading{Real numbers as programs}

\vspace*{0.5cm}

However, given our modern perspective, we can say that after all many more
numbers than Kronecker would have accepted {\em do} have a finite
representation.

This is the program used to calculate them to greater and greater precision.

For example, we can write a program that will produce for any argument {\red
$n$} the first {\red $n$} digits of {\red $\pi$}.

Alternatively it can produce a rational number {\red $r$} such that
{\red $|\pi - r| < 2^{-n}$}.

Whatever approach is taken to the successive approximation of a real number,
the key point is that its representation, the program itself, is finite.

\end{rslide}



\begin{rslide}

\heading{Our representation of reals}

\vspace*{0.5cm}

We  represent a real {\red $x$} by a function {\red $f_x:\nat \to \num$} that
for each {\red $n \in \nat$}:
{\red $$ |f_x(n) - 2^n x| < 1 $$}
This is of course equivalent to
{\red $$ |{f_x(n) \over 2^n} - x| < {1 \over 2^n} $$}
We can actually represent the arithmetic operations on numbers as
higher order functions.

Given functions for approximating {\red $x$} and {\red $y$}, will produce new ones for
approximating {\red $x + y$}, {\red $x y$}, {\red $sin(x)$} and so on, for a wide range of
functions.

Such a result is exact, in the sense that we can then give it an arbitrary
desired precision and it will perform the appropriate calculation
automatically.

\end{rslide}



\begin{rslide}

\heading{Arbitrary precision integers}

\vspace*{0.5cm}

CAML's standard integers (type {\black \tt int}) have a severely limited range, so
first of all we need to set up a type of unlimited-precision integers.

The CAML release includes a library that gives a fast implementation of
arbitrary precision integer (and in fact rational) arithmetic.

A version of CAML Light with this library pre-loaded is installed on Thor. Just
run it with:

\begin{black}\begin{verbatim}
  $ camllight my_little_caml
\end{verbatim}\end{black}

then do the following to get access to all the functions:

\begin{black}\begin{verbatim}
  #open "num";;
\end{verbatim}\end{black}

This library sets up a new type {\black \tt num} of arbitrary precision rational
numbers; we will just use the integer subset.

\end{rslide}


\begin{rslide}

\heading{Operators on {\black \tt num}}

\vspace*{0.5cm}

CAML does not provide overloading, so it is necessary to use different symbols
for the usual arithmetic operations over {\black \tt num}.

Note that small constants of type {\black \tt num} must be written as {\black \tt Int k}
rather than simply {\black \tt k}.

The unary negation on type {\black \tt num} is written {\black \tt minus\_num}. Another handy
unary operator is {\black \tt abs\_num}, which finds absolute values, i.e. given {\red $x$}
returns {\red $|x|$}.

The usual complement of binary operations are also available. The (truncating)
division and modulus function, called {\black \tt quo\_num} and {\black \tt mod\_num}, do not
have infix status. However most of the other binary operators are infixes, and
the names are derived from the usual ones by adding a slash `{\black \verb!/!}'.

\end{rslide}



\begin{rslide}

\heading{Binary operators}

\vspace*{0.5cm}

\bigskip
\begin{tabular}{|l|l|l|}
\hline
Op         & Type                             & Meaning               \\
\hline
{\black \tt **/}  & {\black \tt num -> num -> num}          & Exponentiation        \\
{\black \tt */}   & {\black \tt num -> num -> num}          & Multiplication        \\
{\black \tt +/}   & {\black \tt num -> num -> num}          & Addition              \\
{\black \tt -/}   & {\black \tt num -> num -> num}          & Subtraction           \\
{\black \tt =/}   & {\black \tt num -> num -> bool}         & Equality              \\
{\black \tt <>/}  & {\black \tt num -> num -> bool}         & Inequality            \\
{\black \tt </}   & {\black \tt num -> num -> bool}         & Less than             \\
{\black \tt <=/}  & {\black \tt num -> num -> bool}         & Less or equal         \\
{\black \tt >/}   & {\black \tt num -> num -> bool}         & Greater than          \\
{\black \tt >=/}  & {\black \tt num -> num -> bool}         & Greater or equal      \\
\hline
\end{tabular}
\bigskip

\end{rslide}



\begin{rslide}

\heading{Getting started}

\vspace*{0.5cm}

Recall that our real numbers are supposed to be (represented by) functions
{\red $\num \to \num$}.

In ML we will actually use {\black \tt int -> num}, since the inbuilt
type of integers is more than adequate for indexing the level of accuracy.

Now we can define some basic operations on reals. The most basic operation,
which gets us started, is to produce the real number corresponding to an
integer. This is easy:

\begin{black}\begin{verbatim}
  #let real_of_int k n =
     (Int 2 **/ Int n) */ Int k;;
  real_of_int : int -> int -> num = <fun>
  #real_of_int 23;;
  - : int -> num = <fun>
\end{verbatim}\end{black}

Evidently this satisfies the error criterion: in fact the error is zero.

\end{rslide}





\begin{rslide}

\heading{Basic operations}

\vspace*{0.5cm}

Now we can define the first nontrivial operation, that of unary negation:
\begin{black}\begin{verbatim}
  let real_neg f n = minus_num(f n);;
\end{verbatim}\end{black}
The compiler generalizes the type more than intended, but this will not trouble
us. It is almost as easy to see that the approximation criterion is preserved.
If we know that for each {\red $n$}:
{\red $$ |f_x(n) - 2^n x| < 1 $$}
\noindent then we have for any {\red $n$}:
{\red \begin{eqnarray*}
|f_{-x}(n) - 2^n (-x)| & = & |-f_x(n) - 2^n (-x)|       \\
                       & = & |-(f_x(n) - 2^n x)|        \\
                       & = & |f_x(n) - 2^n x|           \\
                       & < & 1
\end{eqnarray*}}
Similarly, we can define an `absolute value' function on real numbers, using
the corresponding function {\black \tt abs\_num} on numbers.

\end{rslide}





\begin{rslide}

\heading{Addition: first attempt}

\vspace*{0.5cm}

We could define:

{\red $$ f_{x + y}(n) = f_x(n) + f_y(n) $$}

However this gives no guarantee that the approximation criterion is maintained;
we would have:

{\red \begin{eqnarray*}
&      & |f_{x + y}(n) - 2^n (x + y)| \\
& =    & |f_x(n) + f_y(n) - 2^n (x + y)|      \\
& \leq & |f_x(n) - 2^n x| + |f_y(n) - 2^n y|
\end{eqnarray*}}

We can guarantee that the sum on the right is less than {\red $2$}, but not that it is
less than {\red $1$} as required. Therefore, we need in this case to evaluate {\red $x$} and
{\red $y$} to {\em greater} accuracy than required in the answer.

\end{rslide}


\begin{rslide}

\heading{Addition: second attempt}

\vspace*{0.5cm}

Suppose we define:
{\red $$ f_{x + y}(n) = (f_x(n + 1) + f_y(n + 1)) / 2 $$}
\noindent Now we have:
{\red \begin{eqnarray*}
&   & |f_{x + y}(n) - 2^n (x + y)|                              \\
& = & |(f_x(n + 1) + f_y(n + 1)) / 2 - 2^n (x + y)|             \\
& \leq & |f_x(n + 1) / 2 - 2^n x| + |f_y(n + 1) / 2 - 2^n y|    \\
& = & {1 \over 2} |f_x(n + 1) - 2^{n + 1} x| +
      {1 \over 2} |f_y(n + 1) - 2^{n + 1} y|                    \\
& < & {1 \over 2} 1 + {1 \over 2} 1 = 1
\end{eqnarray*}}

Apparently this just gives the accuracy required. However we have implicitly
used real mathematical division above. Since the function is supposed to yield
an integer, we are obliged to round the quotient to an integer.

\end{rslide}




\begin{rslide}

\heading{Rounding division}

\vspace*{0.5cm}

If we just use {\black \tt quo\_num}, the error from rounding this might be almost
{\red $1$}, after which we could never guarantee the bound we want, however accurately
we evaluate the arguments.

We need a division function that always returns the integer closest to the true
result (or one of them in the case of two equally close ones), so that the
rounding error never exceeds {\red $1 \over 2$}.

\begin{black}\begin{verbatim}
  #let ndiv x y = round_num(x // y);;
  ndiv : num -> num -> num = <fun>
  ##infix "ndiv";;
  #(Int 23) ndiv (Int 5);;
  - : num = Int 5
  #(Int 22) ndiv (Int 5);;
  - : num = Int 4
\end{verbatim}\end{black}

Now we are ready to define a correct addition function!

\end{rslide}


\begin{rslide}

\heading{Addition: third attempt}

\vspace*{0.5cm}

Now if we define:
{\red $$ f_{x + y}(n) = (f_x(n + 2) + f_y(n + 2)) \mbox{ ndiv } 4 $$}
\noindent everything works:
{\red \begin{eqnarray*}
&      & |f_{x + y}(n) - 2^n (x + y)|                                   \\
& =    & |((f_x(n + 2) + f_y(n + 2)) \mbox{ ndiv } 4) - 2^n (x + y)|    \\
& \leq & {1 \over 2} + |(f_x(n + 2) + f_y(n + 2)) / 4 - 2^n (x + y)|    \\
& =    & {1 \over 2} + {1 \over 4} |(f_x(n + 2) + f_y(n + 2)) -
                                     2^{n + 2}(x + y)|                  \\
& \leq & {1 \over 2} + {1 \over 4} |f_x(n + 2) - 2^{n + 2} x| +         \\
&      &               {1 \over 4} |f_y(n + 2) - 2^{n + 2} y|           \\
& <    & {1 \over 2} + {1 \over 4} 1 + {1 \over 4} 1 = 1
\end{eqnarray*}}
\noindent Accordingly we make our definition:
\begin{black}\begin{verbatim}
  let real_add f g n =
    (f(n + 2) +/ g(n + 2)) ndiv (Int 4);;
\end{verbatim}\end{black}

\end{rslide}




\begin{rslide}

\heading{Multiplication by an integer (1)}

\vspace*{0.5cm}

It's worth treating this special case efficiently.
We define:
{\red $$ f_{m x}(n) = (m f_x(n+p+1)) \mbox{ ndiv } 2^{p+1} $$}
\noindent where {\red $p$} is chosen so that {\red $2^p \geq |m|$}. For correctness, we have:

{\red \begin{eqnarray*}
&      & |f_{m x}(n) - 2^n (m x)|                                       \\
& \leq & {1 \over 2} + |{m f_x(n+p+1) \over 2^{p+1}} - 2^n (m x)|       \\
& =    & {1 \over 2} + {|m| \over 2^{p+1}} |f_x(n+p+1) - 2^{n+p+1} x|   \\
& <    & {1 \over 2} + {|m| \over 2^{p+1}}                              \\
& \leq & {1 \over 2} + {1 \over 2} {|m| \over 2^p}                      \\
& \leq & {1 \over 2} + {1 \over 2} = 1
\end{eqnarray*}}

\end{rslide}


\begin{rslide}

\heading{Multiplication by an integer (2)}

\vspace*{0.5cm}

In order to implement this, we need a function to find the appropriate {\red $p$}:
\begin{black}\begin{verbatim}
let log2 =
  let rec log2 x y =
    if x </ Int 1 then y
    else log2 (quo_num x (Int 2)) (y + 1) in
  fun x -> log2 (x -/ Int 1) 0;;
\end{verbatim}\end{black}
\noindent The implementation is simply:
\begin{black}\begin{verbatim}
let real_intmul m x n =
  let p = log2 (abs_num m) in
  let p1 = p + 1 in
  (m */ x(n + p1)) ndiv (Int 2 **/ Int p1);;
\end{verbatim}\end{black}
\noindent For division by an integer, we define:
{\red $$ f_{x / m}(n) = f_x(n) \mbox{ ndiv } m $$}

\end{rslide}




\begin{rslide}

\heading{General multiplication}

\vspace*{0.5cm}

This is harder because the error in approximation for one number is multiplied
by the magnitude of the second number. We need a first evaluation to determine
their approximate magnitude.

\begin{black}\begin{verbatim}
  let real_mul x y n =
    let n2 = n + 2 in
    let r = n2 / 2 in
    let s = n2 - r in
    let xr = x(r)
    and ys = y(s) in
    let p = log2 xr
    and q = log2 ys in
    if p = 0 & q = 0 then Int 0 else
    let k = q + r + 1
    and l = p + s + 1
    and m = p + q + 4 in
    (x(k) */ y(l)) ndiv (Int 2 **/ Int m);;
\end{verbatim}\end{black}

The error analysis is long but not really difficult.

\end{rslide}





\begin{rslide}

\heading{Multiplicative inverse (1)}

\vspace*{0.5cm}

In order to get any sort of upper bound on the inverse, let alone a good
approximation, we need to get a {\em lower} bound for the argument.

So first define the {\black \tt msd} function:

\begin{black}\begin{verbatim}
  let msd =
    let rec msd n x =
      if abs_num(x(n)) >/ Int 1 then n
      else msd (n + 1) x in
    msd 0;;
\end{verbatim}\end{black}

This finds the first {\red $n$} with {\red $|f_x(n) | > 1$}.

Note that if {\red $x = 0$} then this will loop indefinitely. This problem is
inevitable since equality of reals, like equality of functions, is not
computable.

However if {\red $x \not= 0$} the above must terminate eventually.

\end{rslide}



\begin{rslide}

\heading{Multiplicative inverse (2)}

\vspace*{0.5cm}

Now we can code the inverse.

\begin{black}\begin{verbatim}
  let real_inv x n =
    let x0 = x(0) in
    let k =
      if x0 >/ Int 1 then
        let r = log2 x0 - 1 in
        let k0 = n + 1 - 2 * r in
        if k0 < 0 then 0 else k0
      else
        let p = msd x in
        n + 2 * p + 1 in
    (Int 2 **/ Int (n + k)) ndiv (x(k));;
\end{verbatim}\end{black}

Now, of course, it is straightforward to define division:

\begin{black}\begin{verbatim}
  let real_div x y =
    real_mul x (real_inv y);;
\end{verbatim}\end{black}

\end{rslide}



\begin{rslide}

\heading{Ordering and equality}

\vspace*{0.5cm}

To decide the ordering relation of {\red $x$} and {\red $y$} it suffices to
find an {\red $n$} such that {\red $|f_x(n) - f_y(n)| \geq 2$}. For example, if
{\red $f_x(n) \geq f_y(n) + 2$} we have
{\red $$ 2^n x > f_x(n) - 1 \geq f_y(n) + 1 > 2^n y $$}
\noindent and so {\red $x > y$}.

\begin{black}\begin{verbatim}
  let rec separate n x y =
    let d = x(n) -/ y(n) in
    if abs_num(d) >/ Int 1 then d
    else separate (n + 1) x y;;
  let real_gt x y = separate x y >/ Int 0;;
  let real_ge x y = real_gt x y;;
  let real_lt x y = separate x y </ Int 0;;
  let real_le x y = real_lt x y;;
\end{verbatim}\end{black}
Note that the only way to arrive at the reflexive orderings is to justify the
irreflexive version. In general, all these will fail to terminate if {\red $x = y$}.

\end{rslide}



\begin{rslide}

\heading{Testing}

\vspace*{0.2cm}

We first define a function to show us a real:
\begin{black}\begin{verbatim}
  let view x d =
    let n = 4 * d in
    let out = x(n) // (Int 2 **/ Int n) in
    approx_num_fix d out;;
\end{verbatim}\end{black}
Now we can test out some simple examples:
\begin{black}\begin{verbatim}
  #let x = real_of_int 3;;
  x : int -> num = <fun>
  #let xi = real_inv x;;
  xi : int -> num = <fun>
  #let wun = real_mul x xi;;
  wun : int -> num = <fun>
  #view x 20;;
  it : string = "3.00000000000000000000"
  #view xi 20;;
  it : string = ".33333333333333333333"
  #view wun 20;;
  it : string = "1.00000000000000000000"
\end{verbatim}\end{black}

\end{rslide}





\begin{rslide}

\heading{The problem of reevaluation}

\vspace*{0.5cm}

The following is very slow:

\begin{black}\begin{verbatim}
  #let x1 = real_of_int 1 in
   let x2 = real_mul x1 x1 in
   let x3 = real_mul x2 x2 in
   let x4 = real_mul x3 x3 in
   let x5 = real_mul x4 x4 in
   let x6 = real_mul x5 x5 in
   let x7 = real_mul x6 x6 in
   view x7 10;;
   - : string = "+1.0000000000"
\end{verbatim}\end{black}

The problem is the repeated evaluation of the same expression. This is worse
because multiplication and inversion often evaluate their arguments more than
once. This tends to lead to a blowup exponential in the depth of the
expression.

\end{rslide}



\begin{rslide}

\heading{Caching}

\vspace*{0.5cm}

We can solve the problem by giving each function a `memo'.

It will remember the most accurate result it has already calculated. If asked
for the same one, or a less accurate one, it can take it from the store.

If we know that {\red $|f_x(n + k) - 2^{n + k} x| < 1$} then we have:
{\red \begin{eqnarray*}
&      & |f_x(n + k) \mbox{ ndiv } 2^ k - 2^n x|                        \\
& \leq & {1 \over 2} + |{f_x(n + k) \over 2^k} - 2^n x|                 \\
& =    & {1 \over 2} + {1 \over 2^k} |f_x(n + k) - 2^{n + k} x|         \\
& <    & {1 \over 2} + {1 \over 2^k}                                    \\
& \leq & 1
\end{eqnarray*}}

Hence we are always safe in returning {\red $f_x(n + k) \mbox{ ndiv } 2^k$}.

\end{rslide}



\begin{rslide}

\heading{Memo function}

\vspace*{0.5cm}

We can use the following generic function to attach a memo to a function:
\begin{black}\begin{verbatim}
let memo f =
  let mem = ref (-1,Int 0) in
  fun n ->
    let (m,res) = !mem in
    if n <= m then
      if m = n then res
      else res ndiv (Int 2 **/ Int(m - n))
    else
      let res = f n in
      mem := (n,res); res;;
\end{verbatim}\end{black}
Now we systematically insert this in each operator. The resulting system is
much more efficient.
\begin{black}\begin{verbatim}
let real_add f g = memo (fun n ->
   (f(n + 2) +/ g(n + 2)) ndiv (Int 4));;
...
\end{verbatim}\end{black}

\end{rslide}




\end{document}

%%% Local Variables:
%%% TeX-master: "all"
%%% End:
